---
title: "1.2 Discrete statistical distributions"
author: "Jan SmyÄka"
date: "17/01/2024"
output: ioslides_presentation
---

## Bernoulli distribution

Tossing a coin

```{r setup, echo=FALSE, warning=F, error=F, message=F}
library(Rlab)
set.seed(113)
```

```{r bernoulli1}
rbern(1, p=0.5)
```

## Bernoulli distribution

The coin is biased

```{r bernoulli2}
rbern(1, p=0.7)
```

## Bernoulli distribution

Tossing a coin multiple times

```{r bernoulli3}
bern=rbern(100, p=0.7)
bern
```

## Bernoulli distribution

The mean expected result equals to p
```{r bernoulli4}
mean(bern)
```

## Bernoulli distribution

The mean result equals to p
```{r bernoulli5}
hist(bern)
```

## Bernoulli distribution

Application: Did my experimental seed germinate?

## Binomial distribution

I toss a coin multiple times and I am interested in the sum of results

```{r binom1}
rbinom(1, size=10, p=0.7)

binom=rbinom(100, size=10, p=0.7)
binom

```

## Binomial distribution

The mean expected result equals to size*p

```{r binom2}
10*0.7
mean(binom)
```

## Binomial distribution
```{r binom3}
hist(binom)
```

## Binomial distribution

Application: I have sown 100 seeds. How many of them germinated?

## Exercise 1

Simulate a sample from a binomial distribution of size=20 and p=0.3 using only rbern() function.

## Exercise 1
```{r binom4}
ex1=rbern(20, 0.3)
sum(ex1)
```

## Poisson distribution
I toss a coin $\infty$ times. The mean expected result is size*p. But the size is $\infty$

$\infty$ * p = $\infty$ for any p>0

$\infty$ * 0 = $\lambda$ , where $\lambda$>0

-> lets use $\lambda$ for controlling the distribution, rather than p

```{r poiss1}
pois=rpois(100, lambda=4)
pois
```

## Poisson distribution
The mean expected result equals to $\lambda$ 
```{r poiss2}
mean(pois)
```
... of course that's how we defined $\lambda$ 

## Poisson distribution
```{r poiss3}
hist(pois)
```

## Poisson distribution

Application: How many pollinators visited the observed plant? The number of potential visitors is theoretically unlimited. 

## Exercise 2
Simulate 1000 samples from a distribution that is reasonably close to Poisson with $\lambda$=3, using only rbinom() function.

## Exercise 2
```{r poiss4}
approxpois=rbinom(1000, size=3000000, p=0.000001)
exactpois=rpois(1000, lambda = 3)
par(mfrow=c(1,2)); hist(approxpois); hist(exactpois)
par(mfrow=c(1,1))

```

