---
title: "2.1 Doing it Bayesian"
author: "Jan Smyčka"
date: "09/01/2024"
output: ioslides_presentation
---

## Frequentist statistics

Data are probabilistic! 

How probable it is to get these interesting data if nothing interesting is going on?

$$p\ value = P(data|null \ hypothesis)$$

Improbable? Something interesting is going on!

## Frequentist statistics

Data are probabilistic! 

How probable are the data given hypothesis x? How likely is the hypothesis x given the data?

$$L(hypothesis \ x|data) = P(data|hypothesis\ x)$$

Lets find the most likely hypothesis based on the data!

## Bayesian statistics

Hypotheses are also probabilistic! How probable are different hypotheses based on data?

$$P(hypothesis|data)$$

(We use here probability as a measure of belief.)

## A bit of algebra again

Rule for joint (AND) probability is
$$P(A \cap B) = P(A) \times P(B|A)$$
$A$ and $B$ can be swapped arbitrarily
$$P(A \cap B) = P(B) \times P(A|B)$$

## A bit of algebra again
and so
$$P(B) \times P(A|B) = P(A) \times P(B|A)$$
which we can rearrange to get
$$P(A|B) = \frac {P(B|A) \times P(A)}{P(B)}$$
which is **the Bayes rule**.

## Conditional probability
We can also say that
$$P(B)=P(B|A) \times P(A)+P(B|nonA) \times P(nonA)$$
so
$$P(A|B) =\frac {P(B|A) \times P(A)}{P(B)}$$

$$=\frac {P(B|A) \times P(A)}{P(B|A) \times P(A)+P(B|nonA) \times P(nonA)}$$

## Conditional probability
if we use sum sign
$$P(A|B)=\frac {P(B|A) \times P(A)}{\sum_A P(B|A) \times P(A)}$$

## Bayes rule in statistics
we can replace $A$ and $B$ by model parameters $\theta$ (or $H$) and the data $x$ (or $d$) to get 

$p(\theta|x) = \frac {p(\theta) \times p(x|\theta)}{p(x)}$

where

$p(x|\theta)$ ... likelihood

$p(\theta)$ ... prior

$p(\theta|x)$ ... posterior

$p(x)$ ... the horrible thing

## Why is p(x) horrible?

$$p(x)=\sum_\theta p(\theta) \times p(x|\theta)$$

$$p(x)=\int_\theta p(\theta) \times p(x|\theta) d\theta$$

## The data

We have five data points from poisson distribution

```{r}
x=rpois(5,20)
x
```

We would like to know posterior distribution of $\lambda$

## Analytical solution

Poisson distribution posterior can be expressed as

$$p(\lambda|x)=\dfrac{p(x|\lambda)*p(\lambda)}{p(x)}$$

where

$$p(x|\lambda)=\prod\limits_{i=1}^n \dfrac{\lambda^{x_i}e^{-\lambda}}{x_i!}; p(\lambda)=0.01\\$$


$$p(x)=\int_\lambda (\prod\limits_{i=1}^n \dfrac{\lambda^{x_i}e^{-\lambda}}{x_i!}*0.01)  d\lambda\\ $$

## Analytical solution - properties

- is accurate
 
- cannot be obtained for most model types


## Numerical computation
We can try to compute numerically

The numerator can be expressed as R function

```{r}
baypo=function(l){
  #likelihood
  lh=function(l){
    prod(((l^x)*exp(-l))/factorial(x))
  }
  #likelihood and prior
  100000000*lh(l)*dunif(l,0,100)
  }
```


## Numerical computation

Now we can generate a grid (or whatever else) and feed the function
```{r}
lambda=seq(0,100,0.01)
postdist=NULL
for (i in 1:length(lambda)){
  postdist[i]=baypo(lambda[i])
  }
prob=postdist/sum(postdist)

```

## Numerical computation

```{r}
plot(prob~lambda, type="l")

```

## Numerical computation - properties

 - is terribly slow and wasteful

 - theoretically works always
 
 - can be paralelized on clusters
